---
title: "Group Project 1: Group 2"
author: "Alyssa Bautista, Raquel Gonzalez, Celia Lopez, and Quynh Pham"
date: "2026-03-08"
output: html_document
---

## Data Management & Wrangling

### Data Import

```{r}
# Import the dataset using a relative path
hints_data <- read.csv("data/HINTSData_2020_clean.csv")
# Preview the data to ensure it loaded correctly
head(hints_data)
```

### Data Cleaning

```{r}
# Check for missing values (NAs) by nesting the sum() and is.na() functions
sum(is.na(hints_data))

# Factor the categorical variables
# Extract and replace the Quality Care data and use the factor() function to assign levels and labels to the data
hints_data$QualityCare <- factor(hints_data$QualityCare, levels = c(1, 2, 3, 4, 5), labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"), ordered = TRUE)

# Extract and replace the BirthGender data and use the factor() function to assign levels and labels to the data
hints_data$BirthGender <- factor(hints_data$BirthGender, levels = c(1, 2), labels = c("Male", "Female"), ordered = TRUE)

# Extract and replace the smokeStat data and use the factor() function to assign levels and labels to the data
hints_data$smokeStat <- factor(hints_data$smokeStat, levels = c(1, 2, 3), labels = c("Current", "Former", "Never"), ordered = TRUE)

# Extract and replace the RaceEthn5 data and use the factor() function to assign levels and labels to the data
hints_data$RaceEthn5 <- factor(hints_data$RaceEthn5, levels = c(1, 2, 3, 4, 5), labels = c("Non-Hispanic White", "Non-Hispanic Black", "Hispanic", "Non-Hispanic Asian", "Non-Hispanic Other"), ordered = TRUE)

```

### Extracting Relevant Variables

```{r}
# Use the subset() function to create a new dataframe consisting of selected variables from the original HINTS dataset

hints_clndata <- subset(hints_data, select = c(HHID, PersonID, Age, AvgDrinksPerWeek, WeeklyMinutesModerateExercise, BMI, QualityCare, BirthGender, smokeStat, RaceEthn5))
```

## Univariate Visualization

### Base R Visualization

```{r}
# Create a bar plot for Quality Care 
qc <- table(hints_clndata$QualityCare)
barplot(qc,
        main = "Distribution of QualityCare",
        xlab = "QualityCare",
        ylab = "Count")
# Display the bar plot
qc
```

```{r}
# Create a RaceEthn5 pie chart
race <- table(hints_clndata$RaceEthn5)
pie(race,
    main = "Distribution of RaceEthn5")
# Display pie chart
race
```

### ggplot2 Visualization

```{r}
# Load packages
library(ggplot2)
```

Using ggplot2 to create charts depicting the distributions of the QualityCare and RaceEthn5 variables.

```{r}
# Use ggplot2 to create a bar plot to display the QualityCare data
qualcare_bar <- ggplot(hints_clndata, aes(x = QualityCare)) +
  geom_bar(fill = "hotpink4", color = "palevioletred1") +
  ggtitle("Quality of Care Distribution") + 
  # Center the title
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Quality of Care") +
  ylab("Frequency")
# Display the chart
qualcare_bar
```

```{r}
# Use ggplot2 to create a lollipop chart to display the RaceEthn5 data
race_ethn_pop <- ggplot(hints_clndata, aes(x = RaceEthn5)) +
  # Create the "lollipop stem"
  geom_segment(aes(xend = RaceEthn5, yend = after_stat(count)), stat = "count", y = 0, color = "springgreen4") +
  # Create the "lollipop head"
  geom_point(stat = "count", size = 6, color = "palegreen1") +
  ggtitle("Lollipop Chart of Ethnic Distributions") +
  # Center the title
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Ethnicities") +
  ylab("Frequency")
# Display the chart
race_ethn_pop
```

Interpretations: The majority of participants were Non-Hispanic White, followed by Hispanic. Overall, participants primarily received a "very good" quality of care, followed by "excellent" and "good".


Using ggplot2 to create charts depicting the distributions of the Age and BMI variables.

```{r}
# Use ggplot2 to create a histogram to display the age data

age_hist <- ggplot(hints_clndata, aes(x = Age)) +
  # Set binwidth to 5 to manage the distribution of data and assign color and borders to the bins.
  geom_histogram(binwidth = 5, fill = "thistle3", color = "mediumorchid4") + 
  ggtitle("Age Distribution") +
  # Center the title
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Age") + 
  ylab("Frequency")
# Display the histogram
age_hist 
```

```{r}
# Use ggplot2 to create a density plot to display the BMI distribution

bmi_dens <- ggplot(hints_clndata, aes(x = BMI)) +
  # Assign a color to fill the chart and the color for the outline. Set opacity to 0.5, so you can see the chart markings
  geom_density(color = "darkcyan", fill = "powderblue", alpha = 0.5) + 
  ggtitle("Density Plot of BMI Distribution") +
  # Center the title
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("BMI") + 
  ylab("Density")
# Display the density plot
bmi_dens 
```

Interpretations: The majority of participants were clustered around the 50-75 age group. The histogram is slightly skewed to the right, but there is a wide distribution of ages, mostly 75 and younger. The majority of participants had a BMI at about 25, and all BMI values generally lied between 20-40. The BMI distribution is right skewed, with values above 40 being outliers. 

### Univariate Visualization Function

Create a single R function that creates univariate distributions for QualityCare and RaceEthn5

```{r}
my_visual_func_I <- function(hints_clndata){
  # Create a table to create a vector of values as input, opposed to raw categorical data
  care_counts <- table(hints_clndata$QualityCare)
  care_plot <- barplot(care_counts, main = "Quality of Care Distribution", xlab = "Quality of Care", ylab = "Frequency", col = "orchid4", border = "mediumorchid4")
  # Display the bar plot
  print(care_plot)
  # Create a new window, so both plots are displayed
  dev.new()
  # Create a table to create a vector of values for the pie chart percentage creation
  ethn_counts <- table(hints_clndata$RaceEthn5)
  # Create the percentages for the pie chart
  ethn_percent <- round(100 * ethn_counts/sum(ethn_counts), 1) 
  # Create a vector of colors for each slice of the pie chart
  my_colors <- c("mediumaquamarine","mediumturquoise", "lightskyblue","steelblue1","midnightblue")
  ethn_plot <- pie(ethn_counts, labels = ethn_percent, main = "Pie Chart of Ethnic Distribution", col = my_colors, radius = 0.8)
  # Shift legend slightly and remove the border around it so it doesn't block the pie chart
  legend("topright", inset = c(-0.1, 0), legend = names(ethn_counts), cex = 0.8, fill = my_colors, bty = "n")
 # Display the pie chart
  print(ethn_plot) 
}
# Execute the function using the HINTS dataset
my_visual_func_I(hints_clndata)
```

Create a single R function that creates univariate distributions for Age and BMI

```{r}
# Create chart labels, assign colors to fill and borders, and create 10 breaks for the histogram distribution.
my_visual_func_II <- function(hints_clndata) {
  hist(hints_clndata$Age, main = "Histogram of Age Distribution", xlab = "Age (Years)", ylab = "Frequency", col = "thistle", border = "mediumorchid4", breaks = 10)
# Create a new window to display both charts
  dev.new()
# Create chart labels and assign colors to fill and the border of the plot using the polygon() function
  bmi_density <- density(hints_clndata$BMI, main = "BMI Density Plot")
  plot(bmi_density, main = "BMI Density Plot")
  polygon(bmi_density, col = "powderblue", border = "darkcyan")
# Display the density plot
  print(bmi_density)
}
# Execute the function using the HINTS dataset
my_visual_func_II(hints_clndata)
```

