---
title: "Group Project 1: Group 2"
author: "Alyssa Bautista, Raquel Gonzalez, Celia Lopez, and Quynh Pham"
date: "2026-03-08"
output: html_document
---

## Data Management & Wrangling

### Data Import

```{r}
# Import the dataset using a relative path
hints_data <- read.csv("data/HINTSData_2020_clean.csv")
# Preview the data to ensure it loaded correctly
head(hints_data)
```

### Data Cleaning

```{r}
# Check for missing values (NAs) by nesting the sum() and is.na() functions
sum(is.na(hints_data))

# Factor the categorical variables
# Extract and replace the Quality Care data and use the factor() function to assign levels and labels to the data
hints_data$QualityCare <- factor(hints_data$QualityCare, levels = c(1, 2, 3, 4, 5), labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"), ordered = TRUE)

# Extract and replace the BirthGender data and use the factor() function to assign levels and labels to the data
hints_data$BirthGender <- factor(hints_data$BirthGender, levels = c(1, 2), labels = c("Male", "Female"), ordered = TRUE)

# Extract and replace the smokeStat data and use the factor() function to assign levels and labels to the data
hints_data$smokeStat <- factor(hints_data$smokeStat, levels = c(1, 2, 3), labels = c("Current", "Former", "Never"), ordered = TRUE)

# Extract and replace the RaceEthn5 data and use the factor() function to assign levels and labels to the data
hints_data$RaceEthn5 <- factor(hints_data$RaceEthn5, levels = c(1, 2, 3, 4, 5), labels = c("Non-Hispanic White", "Non-Hispanic Black", "Hispanic", "Non-Hispanic Asian", "Non-Hispanic Other"), ordered = TRUE)

```

### Extracting Relevant Variables

```{r}
# Use the subset() function to create a new dataframe consisting of selected variables from the original HINTS dataset

hints_clndata <- subset(hints_data, select = c(HHID, PersonID, Age, AvgDrinksPerWeek, WeeklyMinutesModerateExercise, BMI, QualityCare, BirthGender, smokeStat, RaceEthn5))
```


## Descriptive Analysis & Function Writing 

### Statistical Summaries
```{r}
library(dplyr)
#load packages
hints <- read.csv("data/HINTSData_2020_clean.csv")
#import data
head(hints)
#view first row
cont_vars <- hints %>%
  select(Age, AvgDrinksPerWeek, WeeklyMinutesModerateExercise, BMI)
#select only continuous variables
summary(cont_vars)
#calculate 5-number summary
mean(hints$Age, na.rm = TRUE)
mean(hints$AvgDrinksPerWeek, na.rm = TRUE)
mean(hints$WeeklyMinutesModerateExercise, na.rm = TRUE)
mean(hints$BMI, na.rm = TRUE)
#calculate mean
sd(hints$Age, na.rm = TRUE)
sd(hints$AvgDrinksPerWeek, na.rm = TRUE)
sd(hints$WeeklyMinutesModerateExercise, na.rm = TRUE)
sd(hints$BMI, na.rm = TRUE)
#calculate standard deviation
var(hints$Age, na.rm = TRUE)
var(hints$AvgDrinksPerWeek, na.rm = TRUE)
var(hints$WeeklyMinutesModerateExercise, na.rm = TRUE)
var(hints$BMI, na.rm = TRUE)
#calculate variance
#interpretation: the average age of respondents was 55.6 years, with a standard deviation of 16.58, which indicates that the sample includes many middle-aged and older adults. Participants reported an average of 3.36 alcoholic drinks per week, with a median of 0 drinks, suggesting that many respondents don’t drink alcohol. The mean weekly moderate exercise was 161.8 minutes, but the large standard deviation showed high variability in exercise levels. The average BMI was 28.61, which falls in the overweight range, suggesting many participants may be at risk for weight-related health conditions.

```

```{r}


#Use attach function for more efficient coding
attach(hints_clndata)

```


```{r}

#Calculate the summary, mean, standard deviation, and variance of the AGE variable 
summary(Age)


mean(Age)


sd(Age)


var(Age)


#The age range for the sample population in the dataset ranges from 18 to 100 years old, with the average mean age being about 55.59. The standard deviation of the age variable is 16.57902. Squaring the standard deviation gives a variance of 274.864.


```

```{r}
#Calculate the summary, mean, standard deviation, and variance of the AVGDRINKSPERWEEK variable 
summary(AvgDrinksPerWeek)


mean(AvgDrinksPerWeek)


sd(AvgDrinksPerWeek)


var(AvgDrinksPerWeek)


#The average mean amount of alcoholic drinks consumed by the sample population was about 3. The most drinks consumed by one person on average per week was 70, while the majority of the sample had 0 drinks on average per week.

```



```{r}

#Calculate the summary, mean, standard deviation, and variance of the WEEKLYMINUTESMODERATEEXERCISE variable 
summary(WeeklyMinutesModerateExercise)


mean(WeeklyMinutesModerateExercise)


sd(WeeklyMinutesModerateExercise)


var(WeeklyMinutesModerateExercise)


#While about a quarter of the sample population did not exercise moderately at all throughout the week, others were training for upwards of 4620 minutes, or 77 hours! Though not everyone trained so hard, as the average mean weekly minutes of moderate exercise was 161.8, or about 2.7 hours. The high value of the standard deviation, 271.1893,  indicates a wide range of values throughout the variable. 

```


```{r}

#Calculate the summary, mean, standard deviation, and variance of the BMI variable 
summary(BMI)


mean(BMI)


sd(BMI)


var(BMI)


#The average mean BMI for the sample population was 28.6, the lowest BMI being 10.9 and the highest being 73.8. A BMI of 73.8 is considered morbidly obese, whereas a BMI of 10.9 is considered severely underweight. The majority of people fell into the "healthy" or "overweight" category.   



```
```{r}
#part 3
my_summary_func <- function(data, variable)

  values <- data[[variable]]

  results <- list(
    Minimum = min(values, na.rm = TRUE),
    Q1 = quantile(values, 0.25, na.rm = TRUE),
    Median = median(values, na.rm = TRUE),
    Q3 = quantile(values, 0.75, na.rm = TRUE),
    Maximum = max(values, na.rm = TRUE),
    Mean = mean(values, na.rm = TRUE),
    SD = sd(values, na.rm = TRUE),
    Variance = var(values, na.rm = TRUE)
  )

  return(results)
#writing the functions

```
```{r}
my_summary_func(hints, "Age")

my_summary_func(hints, "AvgDrinksPerWeek")

my_summary_func(hints, "WeeklyMinutesModerateExercise")

my_summary_func(hints, "BMI")
#running the function on variables
```
```{r}
#interpretation:The custom function my_summary_func was created to automate the calculation of descriptive statistics for continuous variables. The function returns the five-number summary, mean, standard deviation, and variance for any selected variable in the data set. This allows the same statistical summaries to be quickly generated for multiple variables such as Age, BMI, alcohol consumption, and exercise levels.
```


### Writing functions


```{r}
#Write a single function to display the results of the statistical summaries in B.1

my_summary_func<- function(x){
  return(summary(x))
}


```


```{r}

#Display function of the statistical summaries in B.1
my_summary_func(Age)


my_summary_func(AvgDrinksPerWeek)


my_summary_func(WeeklyMinutesModerateExercise)


my_summary_func(BMI)


```


## Univariate Visualization

### Base R Visualization

```{r}
# Create a bar plot for Quality Care 
qc <- table(hints_clndata$QualityCare)
barplot(qc,
        main = "Distribution of QualityCare",
        xlab = "QualityCare",
        ylab = "Count")
# Display the bar plot
qc
```

```{r}
# Create a RaceEthn5 pie chart
race <- table(hints_clndata$RaceEthn5)
pie(race,
    main = "Distribution of RaceEthn5")
# Display pie chart
race
```
Interpretations: The bar plot shows how respondents rated the quality of care they received. Most responses fall in the higher quality categories. The pie chart shows the distribution of participants across RaceEthn5 groups. Some groups make up a larger portion of the sample than others.

```{r}
# Create the age histogram
hist(hints_clndata$Age,
     main = "Distribution of Age",
     xlab = "Age",
     ylab = "Frequency")
```

```{r}
# Create the BMI boxplot
boxplot(hints_clndata$BMI,
        main = "Distribution of BMI",
        ylab = "BMI")
```

Interpretations: The histogram shows the distribution of age in the dataset. Most participants fall within the middle age range. The boxplot shows the spread of BMI values. A few points appear outside the main range, which may indicate outliers.

### ggplot2 Visualization

```{r}
# Load packages
library(ggplot2)
```

Using ggplot2 to create charts depicting the distributions of the QualityCare and RaceEthn5 variables.

```{r}
# Use ggplot2 to create a bar plot to display the QualityCare data
qualcare_bar <- ggplot(hints_clndata, aes(x = QualityCare)) +
  geom_bar(fill = "hotpink4", color = "palevioletred1") +
  ggtitle("Quality of Care Distribution") + 
  # Center the title
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Quality of Care") +
  ylab("Frequency")
# Display the chart
qualcare_bar
```

```{r}
# Use ggplot2 to create a lollipop chart to display the RaceEthn5 data
race_ethn_pop <- ggplot(hints_clndata, aes(x = RaceEthn5)) +
  # Create the "lollipop stem"
  geom_segment(aes(xend = RaceEthn5, yend = after_stat(count)), stat = "count", y = 0, color = "springgreen4") +
  # Create the "lollipop head"
  geom_point(stat = "count", size = 6, color = "palegreen1") +
  ggtitle("Lollipop Chart of Ethnic Distributions") +
  # Center the title
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Ethnicities") +
  ylab("Frequency")
# Display the chart
race_ethn_pop
```

Interpretations: The majority of participants were Non-Hispanic White, followed by Hispanic. Overall, participants primarily received a "very good" quality of care, followed by "excellent" and "good".


Using ggplot2 to create charts depicting the distributions of the Age and BMI variables.

```{r}
# Use ggplot2 to create a histogram to display the age data

age_hist <- ggplot(hints_clndata, aes(x = Age)) +
  # Set binwidth to 5 to manage the distribution of data and assign color and borders to the bins.
  geom_histogram(binwidth = 5, fill = "thistle3", color = "mediumorchid4") + 
  ggtitle("Age Distribution") +
  # Center the title
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("Age") + 
  ylab("Frequency")
# Display the histogram
age_hist 
```

```{r}
# Use ggplot2 to create a density plot to display the BMI distribution

bmi_dens <- ggplot(hints_clndata, aes(x = BMI)) +
  # Assign a color to fill the chart and the color for the outline. Set opacity to 0.5, so you can see the chart markings
  geom_density(color = "darkcyan", fill = "powderblue", alpha = 0.5) + 
  ggtitle("Density Plot of BMI Distribution") +
  # Center the title
  theme(plot.title = element_text(hjust = 0.5)) +
  xlab("BMI") + 
  ylab("Density")
# Display the density plot
bmi_dens 
```

Interpretations: The majority of participants were clustered around the 50-75 age group. The histogram is slightly skewed to the right, but there is a wide distribution of ages, mostly 75 and younger. The majority of participants had a BMI at about 25, and all BMI values generally lied between 20-40. The BMI distribution is right skewed, with values above 40 being outliers. 

### Univariate Visualization Function

Create a single R function that creates univariate distributions for QualityCare and RaceEthn5

```{r}
my_visual_func_I <- function(hints_clndata){
  # Create a table to create a vector of values as input, opposed to raw categorical data
  care_counts <- table(hints_clndata$QualityCare)
  care_plot <- barplot(care_counts, main = "Quality of Care Distribution", xlab = "Quality of Care", ylab = "Frequency", col = "orchid4", border = "mediumorchid4")
  # Display the bar plot
  print(care_plot)
  # Create a new window, so both plots are displayed
  dev.new()
  # Create a table to create a vector of values for the pie chart percentage creation
  ethn_counts <- table(hints_clndata$RaceEthn5)
  # Create the percentages for the pie chart
  ethn_percent <- round(100 * ethn_counts/sum(ethn_counts), 1) 
  # Create a vector of colors for each slice of the pie chart
  my_colors <- c("mediumaquamarine","mediumturquoise", "lightskyblue","steelblue1","midnightblue")
  ethn_plot <- pie(ethn_counts, labels = ethn_percent, main = "Pie Chart of Ethnic Distribution", col = my_colors, radius = 0.8)
  # Shift legend slightly and remove the border around it so it doesn't block the pie chart
  legend("topright", inset = c(-0.1, 0), legend = names(ethn_counts), cex = 0.8, fill = my_colors, bty = "n")
 # Display the pie chart
  print(ethn_plot) 
}
# Execute the function using the HINTS dataset
my_visual_func_I(hints_clndata)
```

Create a single R function that creates univariate distributions for Age and BMI

```{r}
# Create chart labels, assign colors to fill and borders, and create 10 breaks for the histogram distribution.
my_visual_func_II <- function(hints_clndata) {
  hist(hints_clndata$Age, main = "Histogram of Age Distribution", xlab = "Age (Years)", ylab = "Frequency", col = "thistle", border = "mediumorchid4", breaks = 10)
# Create a new window to display both charts
  dev.new()
# Create chart labels and assign colors to fill and the border of the plot using the polygon() function
  bmi_density <- density(hints_clndata$BMI, main = "BMI Density Plot")
  plot(bmi_density, main = "BMI Density Plot")
  polygon(bmi_density, col = "powderblue", border = "darkcyan")
# Display the density plot
  print(bmi_density)
}
# Execute the function using the HINTS dataset
my_visual_func_II(hints_clndata)
```

